---
title: "Multivariate Regression Analysis"
output: html_document
---


Almost all social phenomena have more than one cause. To control, statistically, for all possible causes social scientists employ multinomial regression analysis.

The multivariate regression model is the following:

<center>

$Y_{i}=\alpha+\beta_{1}X_{i}+ \beta_{2}Z_{i}+ u_{i}$ 

</center>

The interpretation of the slope coefficients for the multivariate- in this case three variable model- is similar to the bivariate model but with one major difference. The coefficient $\beta_{1}$ represents the effect of X on Y while keeping Z (the third variable) constant. The same will hold if we had a model with four variables. 



## Exploring, cleaning, and tansforming your variables

Although many data manipulation function exist in basic R sometimes it is easier to use packages build for the sole purpose of making data manipulation easier. `dplyr` is such a package (https://dplyr.tidyverse.org/) 

`dplyr` is providing a consistent set of verbs that help you solve the most common data manipulation challenges: 


<span style="color:red"> mutate( ) </span> $\rightarrow$  **adds new variables that are functions of existing variables** \
<span style="color:red"> select( ) </span> $\rightarrow$  **picks variables based on their names** \
<span style="color:red"> filter( ) </span> $\rightarrow$   **picks cases based on their values** \
<span style="color:red"> arrange( ) </span> $\rightarrow$   **changes the ordering of the rows** \
<span style="color:red"> summarise( ) </span> $\rightarrow$  **reduces multiple values down to a single summary** \


```{r message=FALSE, warning=FALSE}
library(dplyr)
```




### The %>% operator 

Although not required `dplyr` make use of the pipe (%>%) operator. The main value of the pipe operator is the ability to connect multiple functions together. To give you an example:

<center>

**filter(my_data, my_variable == variable_value)** \
or \
**my_data %>% filter(my_variable == variable_value)** \

</center>


We will use the European Value Survey (EVS) which includes a series of continuous and categorical variables.

### The `select( )` function

`select ()` will keep only those variables in the dataset you are interested in. Sometimes, especially when we re working with big datasets we want to reduce the number of variables. 


In our seminar we will work again with the European Value Study (2019). As you already know the dataset contains many variables.Our goal is to explore what predicts attitudes towards immigration.  The exact wording of the question is the following:


<center>
**Please look at the following statements and indicate where you would place your views on this scale?** 

**Immigrants take jobs** \                             **Immigrants do not take jobs** \ 
**away from [NATIONALITY]** \                            **away from [NATIONALITY]** \ 
                        **1   2   3   4   5   6   7   8   9   10** \

</center>


```{r}
load("EVS_UK_full.RData")
```

Our dependent variable is attitudes towards immigration (v185), our main idependent variables are: self-placement on the left-right spectrum (v102), attitudes towards nationality (v189,v190, v191, v192, v193), and finally three control variables namely age (v226), education (v243_edulvlb), and gender (v225). We also need the variable describing the name of the country in the dataset (country).

```{r message=FALSE, warning=FALSE}
sub.sample<-EVS_UK_full %>% select(v102, v185, v189,v190, v191, v192, v193, v225, v226,v243_edulvlb,country)
head(sub.sample)
```

EVS includes all European countries, for our excercise we want to analyse data examining attitudes towards immigration in GB. To exclude all other countries from our dataset we need a) to know the value of the variable that corresponds to Great Britain (country==826) b) to delete all other values of the country variable from our dataset. To do so we will use the `filter ()` function. 

```{r subsample}
sub.sample %>% filter(country == 826) 
```

Our next step is to let R know which values represent missing cases. According to the codebook all values ranging from $-10$ to $-1$ describe missing cases (Don't know, No answer, not applicable)

```{r missing cases}
sub.sample[sub.sample <=-1] <- NA
sub.sample[sub.sample <=-2] <- NA
```


Our dependent variable is a continuous variable with a range from 1 to 10, where 1 means that the respondent hold anti-immigration attitudes (Immigrants take jobs)  and 10 that the respondent doesn't hold anti-immigration attitudes. When the large number, in our example (10), indicate lower agreement with the phenomenon under study, then we say that the variables are _negatively coded_. This is not really wrong but it makes more sense, is more intuitive, to reverse the order of the coding. It is the case that readers and researchers expect higher values to indicate higher levels of agreement with the phenomenon under study, in this case anti-immigration attitudes.


Using the `arrange( )` function we can reverse the order of the code of the values. In our example what we want is instead of `10` representing positive attitudes towards immigration to represent negative ones.

```{r reverse}
sub.sample<-sub.sample %>% mutate(immi.jobs=(max(v185,na.rm=TRUE)-(v185)))
```
*Note*: I used the na.rm=T function inside the max function `max()`. This is because otherwise due to NAs R cannot make the maths.

To make sure we did everything correctly we use the `table ()` function to compare the two variables- before and after reversing the codes.

```{r compare}
table(sub.sample$immi.jobs)
table(sub.sample$v185)
```

We should apply the same method to reverse the order of three more variables v189 to v193. 

```{r many}
sub.sample<-sub.sample %>% mutate(born.country=(max(v189,na.rm=TRUE)-(v189)))
sub.sample<-sub.sample %>% mutate(respect.inst=(max(v190,na.rm=TRUE)-(v190)))
sub.sample<-sub.sample %>% mutate(country.ancestry=(max(v191,na.rm=TRUE)-(v191)))
sub.sample<-sub.sample %>% mutate(speak.lang=(max(v192,na.rm=TRUE)-(v192)))
sub.sample<-sub.sample %>% mutate(share.cultr=(max(v193,na.rm=TRUE)-(v193)))

head(sub.sample)
```


There is a more efficient way and less time consuming to do the same think as above. This is by using loops.


Our next step is to create the `age` variable, we know the year of birth and that the survey administrated in 2017. To calculate the responder's name when the survey took place we substract respondents year of birth from the year the survey took place (2017- year of birth).  

```{r}
sub.sample<-sub.sample %>% mutate(age=2017-v226)
table(sub.sample$age)
```

Our last step is to use the `rename( )`function, part of `dplyr` to give meaningful names to our variables.

```{r rename}
sub.sample<-sub.sample %>% rename(education=v243_edulvlb)
sub.sample<-sub.sample %>% rename(gender=v225)
sub.sample<-sub.sample %>% rename(left_right=v102)
head(sub.sample)
```


We are now ready to conduct our analysis. We will use the `lm()` function.

```{r mregression}
model.1<-lm(immi.jobs~ born.country+respect.inst+country.ancestry+speak.lang+share.cultr+left_right+education+gender+age,data=sub.sample)
summary(model.1)
```

```{r arm, message=FALSE, warning=FALSE}
library(dotwhisker)
```

```{r}
dwplot(list(model.1), vline = geom_vline(xintercept = 0, colour = "grey60", linetype = 2)) %>% # plot line at zero _behind_ coefs
    relabel_predictors(c(immi.jobs = "Immigrants take jobs",                       
                         born.country = "Born in country", 
                         respect.inst = "Respect institutions", 
                         country.ancestry = "Country ancestry", 
                         speak.lang = "Speak language", 
                         share.cultr = "Share culture",
                         left_right="Left-Right",
                         education="Education",
                         gender="Gender",
                         age="Age")) +
     theme_classic() + xlab("Coefficient Estimate") + ylab("") +
     geom_vline(xintercept = 0, colour = "grey60", linetype = 2) +
     ggtitle("Predicting anti-immigration attitudes") +
     theme(plot.title = element_text(face="bold"),
           legend.position = c(0.01, 0.03),
           legend.justification = c(0, 0), 
           legend.background = element_rect(colour="grey80"),
           legend.title = element_blank()) 

```

From our analysis we can see that two of our variables are not statistical significant, namely age, respect towards institutions, and left-right. Let's construct another model but this time we will exclude the three variables that are not statistical significant. 


```{r}
model.2<-lm(immi.jobs~ born.country+country.ancestry+speak.lang+share.cultr+education+gender,data=sub.sample)
summary(model.2)
```

We can now plot both models and compare the results.

```{r 2plots}
dwplot(list(model.1, model.2), vline = geom_vline(xintercept = 0, colour = "grey60", linetype = 2)) %>% # plot line at zero _behind_ coefs
    relabel_predictors(c(immi.jobs = "Immigrants take jobs",                       
                         born.country = "Born in country", 
                         respect.inst = "Respect institutions", 
                         country.ancestry = "Country ancestry", 
                         speak.lang = "Speak language", 
                         share.cultr = "Share culture",
                         left_right="Left-Right",
                         education="Education",
                         gender="Gender",
                         age="Age")) +
     theme_classic() + xlab("Coefficient Estimate") + ylab("") +
     geom_vline(xintercept = 0, colour = "grey60", linetype = 2) +
     ggtitle("Predicting anti-immigration attitudes") +
     theme(plot.title = element_text(face="bold"),
           legend.position = c(0.01, 0.03),
           legend.justification = c(0, 0), 
           legend.background = element_rect(colour="grey80"),
           legend.title = element_blank()) 
```