---
title: "Binary Dependent Variable"
output: html_document
---

## Loading the data


For this section we will be using data from the British Election Stystem survey to investigate the effect of perceptions about the economy on party choice. This is part of a broader research examining the effect of political issues on party choice and especialy on the effect of the economy on the party in government. 

<a href="https://s3.amazonaws.com/academia.edu.documents/6179351/s4._economic_voting_an_introduction__lewis-beck_paldam_2000_.pdf?response-content-disposition=inline%3B%20filename%3DEconomic_Voting_An_Introduction.pdf&X-Amz-Algorithm=AWS4-HMAC-SHA256&X-Amz-Credential=AKIAIWOWYYGZ2Y53UL3A%2F20191129%2Fus-east-1%2Fs3%2Faws4_request&X-Amz-Date=20191129T110058Z&X-Amz-Expires=3600&X-Amz-SignedHeaders=host&X-Amz-Signature=8a2e30564a1772df23b48cce3f6f246e4582a99d72fb75b66269b96060ff59bd"> Lewis-Beck, Michael S., and Martin Paldam. "Economic voting: an introduction." Electoral studies 19.2-3 (2000): 113-121.</a>

```{r}
load("bes.RData")
```
We will analyse data from the 2017 general election, and we will test whether citizens' who feel that the economy is doing better will support the party in government. 

We can download the full dataset from the website below:

<a href="https://www.britishelectionstudy.com/data-objects/cross-sectional-data/"> British Election Study</a>

We should download the post-election study entitled: 2017 Face-to-face Post-election Survey Version 1.3

We will use the following variables from the dataset: 


Name| Description (name in the dataset in parenthesis)
-------|----------------------------------------------------------------------------
**incumbent**| (1) The respondent will vote for the party in government (0) will vote for another party (b02)
**left-right** | position on the left- right spectrum (e01)
**egocentric_retro**| retrospective evaluation of personal finances (l01)
**egocentric_pros** |  prospective evaluation of personal finances (l02)
**sociotropic_retro** | retrospective evaluation of UK's economy (l04)
**sociotropic_pros**|  prospective evaluation of UK's economy (l03)


## Create a subset  

Our next step is to create a sub-set including only the variables we will include in our analysis

```{r subset, message=F, warning=FALSE}
library(dplyr)
eco_voting<-bes %>%select(b02,Age,edlevel,e01,l01,l02,l03,l04,p01)
View(eco_voting)
```
**Note**: If you use RStudio, you can type the pipe ($%>%$) with **Ctrl + Shift + M** if you have a **PC** or **Cmd + Shift + M** if you have a **Mac**.
## Assign missing cases

Our next step is to replace the negative values (-1, -2) to NAs

```{r na}
eco_voting[eco_voting <=-1] <- NA
eco_voting[eco_voting <=-2] <- NA
```

## Create the dependent variable

Our dependent variable is a dichotomous, dummy variable where $1$ means that the respondent voted for the party in government - this is what we are trying to explain- and $0$ means that they voted for another party. 

```{r dependent}
eco_voting<-eco_voting %>% mutate(incumbent=case_when(b02==2~1, b02==1~0, b02>=3~0)) 
```

You may use the `table()` function to double check the new variable entitled `incumbent':

```{r check}
table(eco_voting$b02)
table(eco_voting$incumbent)
```


## Rename your independent variables

We will use the `rename()` function part of `dplyr`:

```{r rename}
eco_voting<-eco_voting %>% rename(left_right=e01, egocentric_retro=l01,egocentric_pros=l02,sociotropic_pros=l03,sociotropic_retro=l04,brexit=p01)

View(eco_voting)
```


## Logistic Regression Model

We will use the generalized linear model function `glm()` to estimate a logistic regression-remember that we have a dummy dependent variable. The function is very similar to the `lm()` function- the only difference is that there is an additional argument called `family()`. The `family()` function will tell R that we want to estimate a logistic regression. 

Let's see it in practice, all we have to do is to include the following line in the `glm()` function:

```
family = binomial(link = "logit") argument
```


```{r  model.linear}
logit.model<-glm(incumbent~ sociotropic_pros+egocentric_retro+left_right, data=eco_voting,family = binomial(link = "logit"))
summary(logit.model)
```


Interpreting the results of a logistic regression model is not the same as the interpretation of the linear model. Remember that for the linear model the coefficient describe the effect of a unit change (increase or decrease) in X on Y.


For the logistic regression the interpretation of the coefficient is: a one unit change (increase or decrease) in X is associated with a $\hat{\beta}$ change in the log-odds of the dependent variable (Y), holding all other variables constant. 

For example, the coefficient describing  perceptions about the economy `sociotropic_pros` is equal to $0.047$, implying that the log-odds of voting for the party in government are $0.047$ higher when the respondent believe that the economy is doing well, holding all other variables constant.

## Predicted probabilities

As you may already realised log-odds are not straight-forward, this is why we use the `predict()` function which give us predictions for Y, the dependent variable. 

```{r}
probabilities<-predict(logit.model,type = "response")
```


## The effects package

The final section include's the use of the `effects` package developed by Fox, et al. The `effects` package creates plots for various statistical models. In this example I will show you how it works with a logistic model- but it works with linear models, mixed effects models and many other.  

First you have to install the `effects` package. The easiest way to visualise your model is by using the `allEffects()` function in combination with the `plot()` function. 

```{r}
library(effects)
plot(allEffects(logit.model))
```

**References**

Fox, J. (2003) Effect displays in R for generalised linear models. Journal of Statistical Software 8:15, 1–27, http://www.jstatsoft.org/v08/i15/.
